---
title: "QR Factorization"
author: "CaseyJayne"
date: "7/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## QR Factorization using Gram Schmidt Algorithm

 First I wrote the GS and a normalization algorithm.  The R matrix will be made "by hand"

```{r functions}
gs <- function(vectors){
    #'@param vectors a list of vectors
    #'@return a list of orthogonal vectors
    #'@author Casey Jayne

    vecs = list()
    for(i in 1:length(vectors)){
        amount = 0
        for(vals in 1:(i-1)){
            if(length(vecs)>=1){
                t = (t(vecs[[vals]]) %*% vectors[[i]])/crossprod(vecs[[vals]])
                t = t[[1]]*vecs[[vals]]
                amount = amount+t
            }
        }
        vecs[[i]] = vectors[[i]] - amount
    }

    return(vecs)
}

vec_normalize <- function(vectors){
    #'@param vectors a list of vectors
    #'@return a list of normalized vectors by 2norm
    #'@author Casey Jayne
    for(item in 1:length(vectors)){
        vectors[[item]] = vectors[[item]]/sqrt(sum(vectors[[item]]^2)) # standard 2 norm
    }

    return(vectors)
}
```

Create the list of vectors given

```{r given}
m = list(matrix(c(1,0,1)), matrix(c(0,1,1)), matrix(c(1,4,6)))
```

Run the GS Algorithm

```{r gs}
v <- gs(m)
```

Normalize for the Q vectors

```{r q}
# Normalize GS to get the q matrix
q <- vec_normalize(v)
```

Create the R matrix skeleton and put norms of the GS values on diagonal
```{r r_mat}
# Create R matrix with norms of GS on diagonal
r = matrix(0, 3,3)
r[1,1] = sqrt(sum(v[[1]]^2))
r[2,2] = sqrt(sum(v[[2]]^2))
r[3,3] = sqrt(sum(v[[3]]^2))
```

Fill in the upper triangle by inner product of q and original vectors
```{r fillin}
# Fill in R's upper triangle with original vectors cross the normalized GS
r[1,2] = crossprod(q[[1]], m[[2]])
r[1,3] = crossprod(q[[1]], m[[3]])
r[2,3] = crossprod(q[[2]], m[[3]])
```

Put into matrix form to check calculation
```{r QR}
# Put in matrix form to check calculations
Q = matrix(unlist(q), nrow = 3)
R = matrix(unlist(r), nrow = 3)
round(Q %*% R, 2) # equal to original vectors!!
```

*Note the rounding is for an "almost zero" value in position [1,2]*

